### 2023-02-28 calculus.tex
Prefix marking rules with MK

Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
only need the following rules for marking unmarked numbers and typing marked numbers:
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{ISNum}

@@

The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{ISPlus}

@@

Let us now consider the more interesting case of variables. The typing of unmarked variables in the
gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
synthetic rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{ISVar}

@@

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{ISUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing---and the unbound variable to be used in any
expression. % TODO clarify why this is correct

@@

The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
abstraction:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{ISLam}

  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{IALam1}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
the expression does not synthesize a type consistent with the expected type. We then have the
following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{IALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{IALam2}.

A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the expected input type is inconsistent with the actual input
type, giving the last pair of analytic rules for lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{IALam3}

@@

The case of application is somewhat interesting. In the unmarked language, only a synthesis rule is
necessary:
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynType{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology, we can straightforwardly give the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{ISAp1}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. The applied
expression is placed inside the inconsistent type mark, which is in accordance with the meaning of
that mark.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{ISAp2}

@@

It is worth noting, however, that $\EMV_1$ is not actually in analytic position, but in a synthetic
position where any arrow type is expected. Treating such ``constrained synthetic'' cases specially,
we may instead designate a specialized mark, written $\ECApNonMatchedAltAlt{\ECMV}$. This yields the
following alternative rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECApNonMatchedAltAlt{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
  }{ISAp2'}

@@

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ETrue}{\ECTrue}{\TBool}
  }{ISTrue}

@@

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\EFalse}{\ECFalse}{\TBool}
  }{ISFalse}

@@

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
the marked conditional from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{ISIf}

@@

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV} \\
  }{
    \ctxAnaFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{IAIf}

@@

However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}
$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures marking totality on
conditionals:
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
    \inconsistent{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
  }{ISInconsistentBranches}

@@

Since we did not provide explicit analytic rules for marking expressions that are not lambda
abstractions or conditionals (or typing their marked versions) either, we give corresponding
subsumption rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
  }{IASubsume}

@@

However, it may be the case that the expression is inconsistent with the expected type. Recalling
the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyze
$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
rules is required to handle such cases:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\\\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{IAInconsistentTypes}

@@

  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
  \[\begin{array}{rcl}
    & \cdots & \\
    \joinsTo{(\TPair{\TMV_1}{\TMV_2})}{(\TPair{\TMV_1'}{\TMV_2'})}{\TPair{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]

@@

The marked language is extended with the same forms, and we devise the following marking rules,
again modelling off the above unmarked typing rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots
                               \mid \ECPair{\ECMV}{\ECMV}
                               \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                               \mid \ECInconMatchedPair{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{ISPair}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
  }{ISProjL1}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
  }{ISProjR1} \\

  \judgment{
    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{IAPair1}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
  }{IAProjL}

>>>>

Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
only need the following rules for marking unmarked numbers and typing marked numbers:
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{MKSNum}

@@

The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

@@

Let us now consider the more interesting case of variables. The typing of unmarked variables in the
gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
synthetic rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{MKSVar}

@@

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{MKSUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{MKSUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing---and the unbound variable to be used in any
expression. % TODO clarify why this is correct

@@

The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
abstraction:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MKSLam}

  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam1}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
the expression does not synthesize a type consistent with the expected type. We then have the
following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MKALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{MKALam2}.

A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the expected input type is inconsistent with the actual input
type, giving the last pair of analytic rules for lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MKALam3}

@@

The case of application is somewhat interesting. In the unmarked language, only a synthesis rule is
necessary:
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynType{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology, we can straightforwardly give the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{MKSAp1}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. The applied
expression is placed inside the inconsistent type mark, which is in accordance with the meaning of
that mark.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

@@

It is worth noting, however, that $\EMV_1$ is not actually in analytic position, but in a synthetic
position where any arrow type is expected. Treating such ``constrained synthetic'' cases specially,
we may instead designate a specialized mark, written $\ECApNonMatchedAltAlt{\ECMV}$. This yields the
following alternative rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECApNonMatchedAltAlt{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
  }{MKSAp2'}

@@

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ETrue}{\ECTrue}{\TBool}
  }{MKSTrue}

@@

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\EFalse}{\ECFalse}{\TBool}
  }{MKSFalse}

@@

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
the marked conditional from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{MKSIf}

@@

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV} \\
  }{
    \ctxAnaFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKAIf}

@@

However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}
$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures marking totality on
conditionals:
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
    \inconsistent{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
  }{MKSInconsistentBranches}

@@

Since we did not provide explicit analytic rules for marking expressions that are not lambda
abstractions or conditionals (or typing their marked versions) either, we give corresponding
subsumption rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
  }{MKASubsume}

@@

However, it may be the case that the expression is inconsistent with the expected type. Recalling
the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyze
$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
rules is required to handle such cases:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\\\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{MKAInconsistentTypes}

@@

  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    & \cdots & \\
    \joinsTo{(\TPair{\TMV_1}{\TMV_2})}{(\TPair{\TMV_1'}{\TMV_2'})}{\TPair{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]

@@

The marked language is extended with the same forms, and we devise the following marking rules,
again modelling off the above unmarked typing rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots
                               \mid \ECPair{\ECMV}{\ECMV}
                               \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                               \mid \ECInconMatchedPair{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{MKSPair}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
  }{MKSProjL1}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
  }{MKSProjR1} \\

  \judgment{
    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair1}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
  }{MKAProjL}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV}
  }{MKAProjR}
\end{mathpar}
%
accompanied by corresponding typing rules for marked expressions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV_1} \\
    \ctxSynType{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{MSPair}
